---
title: "Assignment 5-ARIMA"
author: "Marleny Tochihuitl-Mendoza"
date: "2025-04-02"
output: html_document
---

```{r}
library(fpp)
library(fpp2)
library(forecast)
library(knitr)
RealPCE <- read.csv("/Users/marleny/Downloads/Rutgers\ 24-25/TimeSeriesModeling/HW4/Consumption.csv", header = TRUE)

names(RealPCE)[1] <- "DATE"
names(RealPCE)[2] <- "Real Personal Consumption Expenditure in USD"
RealPCE$DATE <- as.Date(RealPCE$DATE)
RealPCE$'Real Personal Consumption Expenditure in USD' <- as.numeric(RealPCE$'Real Personal Consumption Expenditure in USD')
kable(RealPCE, caption = "Table: Real Personal Consumption Expenditure in USD")
RealPCE.ts <-ts(RealPCE[,2], start = c(2014,4), end = c(2024,4), frequency = 4)
RealPCE.ts
plot(RealPCE.ts, main = "Real Personal Consumption Expendtiure in the US")
adf.test(RealPCE.ts)
```

#### P-value >0.05 so we need differencing 

```{r}
nsdiffs(RealPCE.ts)
ndiffs(RealPCE.ts)
ndiffs((diff(RealPCE.ts,4)))
RealPCEdiff1 <- diff(RealPCE.ts,lag=4)
tsdisplay(RealPCEdiff1)
# nsdiffs of 1 signifies we need a seasonal differencing of 1. ndiffs is also 1, shows we need one regular difference. However, after applying the seasonal difference no additional regular difference is required.
```

#### To see if more differences are needed I ran the following. The kpss p-value is 0.1, which is greater than 0.05 and adf p-value is 0.01, which is less than 0.05 so no further differencing is needed. Furthermore, Acf only showed 2 significant lags, may benefit from a window to eliminate effects of the pandemic.

```{r}
ndiffs(RealPCEdiff1)
kpss.test(RealPCEdiff1)
adf.test(RealPCEdiff1)
plot(RealPCEdiff1, main = "Time Series using Differencing")
```

### ARIMA Model

```{r}
autofit_PCE <- auto.arima(RealPCE.ts, trace = TRUE, stepwise = FALSE)
autofit_PCE
```


### Residual Analysis
#### Acf: all lags are insiginificant. 
#### Box test: suggests that the residuals are uncorrelated(good)
#### Plot.ts: From 2014 to around 2019 most residuals are near 0. In 2020 the model overpredicts as the residuals are in the -300s, this is due to the global pandemic of COVID-19. After that residuals are between 0 and 200 for a few years until around 2022 in which residuals begin stablizing and centering between around -50 and 50. 
#### Histogram: residuals have a right-skew with a few extreme residuals on the lower end(from the pandemic). However, bulk of the residuals are centered around zero, which is good.

```{r}
Acf(autofit_PCE$residuals)
Box.test(residuals(autofit_PCE), lag=20, type="Ljung")
plot.ts(residuals(autofit_PCE), main = "Residuals of ARIMA model")
hist(autofit_PCE$residuals, main = "Residuals of ARIMA model")
tsdiag(autofit_PCE)
```

### Forecasting and Accuracy

#### MAPE is only 1% which is a good sign that the model has a low percentage error so the forecacst is accurate. 
```{r}
plot(RealPCE.ts, main = "Real Personal Consumption Expendtiure in the US")
lines(autofit_PCE$fitted, col ="red")
forecast_arima.PCE<-forecast(autofit_PCE,h=5)
plot(forecast_arima.PCE)
forecast_arima.PCE
accuracy(forecast_arima.PCE)
```

### ARIMA using window function
```{r}
PCE_window <- window(RealPCE.ts,start=c(2021,1))
PCE_window
plot(PCE_window, main = "Real Personal Consumption Expenditure in the US, starting 2021 Q2")
nsdiffs(PCE_window)
ndiffs(PCE_window)
Acf(PCE_window)
#Acf shows nothing significant
arima_PCEwindow<-auto.arima(PCE_window)
plot(PCE_window, main = "Real Personal Consumption Expenditure in the US, starting 2021 Q2")
lines(arima_PCEwindow$fitted, col = "red")
```

### Residual Analysis 
#### Histogram appears to have more of a bell curve than the previous ARIMA model without the windows function. 
```{r}
Acf(arima_PCEwindow$residuals)
Box.test(residuals(autofit_PCE), lag=20, type="Ljung")
plot.ts(residuals(arima_PCEwindow), main = "Residuals of ARIMA model")
hist(arima_PCEwindow$residuals, main = "Residuals of ARIMA model")
tsdiag(arima_PCEwindow)
```

### Forecasting and Accuracy

```{r}
forecast_window<-forecast(arima_PCEwindow,h=5)
plot(forecast_window)
forecast_window
accuracy(forecast_window)
#MAPE improved to 0.386 rather than 1.04
```
```

